---
title: "Working with 2012-2013 BJS National Crime Victimization Survey Data"
author: "Pauline I. Alvarado"
date: "Fall 2018"
output: pdf_document
---

This coding exercise was from [University of Pennsylvania's Criminal Justice Data Science Course](https://github.com/gregridgeway/R4crim>) taught by Dr. Gregory Ridgeway. Script has been modified from course examples to fit personal notation conventions. Data was taken from the [National Archive of Criminal Justice Data (NACJD)](https://www.icpsr.umich.edu/icpsrweb/NACJD/series/95).  


# Load and view data
## List files
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  # 2012 Files
    list.files("NCVS2012/", recursive = TRUE)

  # 2013 Files
    list.files("NCVS2013/", recursive = TRUE)

```



## Load and categorize data
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  # Address record-type (DS1)
    load("NCVS2012/DS0001/34650-0001-Data.rda")
    load("NCVS2013/DS0001/35164-0001-Data.rda")

    addr12 <- da34650.0001
    addr13 <- da35164.0001
    
  # Household information (DS2)
    load("NCVS2012/DS0002/34650-0002-Data.rda")
    load("NCVS2013/DS0002/35164-0002-Data.rda")

    househld12 <- da34650.0002
    househld13 <- da35164.0002

  # Personal information (DS3)
    load("NCVS2012/DS0003/34650-0003-Data.rda")
    load("NCVS2013/DS0003/35164-0003-Data.rda")

    person12 <- da34650.0003
    person13 <- da35164.0003
    
  # Incident information (DS4)
    load("NCVS2012/DS0004/34650-0004-Data.rda")
    load("NCVS2013/DS0004/35164-0004-Data.rda")

    incident12 <- da34650.0004
    incident13 <- da35164.0004
    
  # Remove objects from the environment
    rm(da34650.0001, da34650.0002, da34650.0003, da34650.0003, da34650.0004, da35164.0001, da35164.0002, da35164.0003, da35164.0004)

```



## View month and year of each incident
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  with(incident12, table(V4014, V4015))
  with(incident13, table(V4014, V4015))
  
```



# Create 2012 incident dataframe
## Merge 2012 and 2013 incident data
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  # Bind rows
    incident2012 <- rbind(incident12, incident13)

  # View years
    table(incident2012$V4015)
    
  # Subset only 2012 information
    incident2012 <- subset(incident2012, V4015 == 2012)
  
```



## Exclude crimes occurring outside US or have unknown location
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  incident2012 <- subset(incident2012, V4022 != "(1) Outside U.S." | is.na(V4022))
  
```



## Only include up to 10 occurrences in a series crime (BJS convention)
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  i <- with(incident2012, which((V4019 == "(2) No (is series)") & (V4016 >= 11) & (V4016 <= 996)))
  incident2012$V4016[i] <- 10
  incident2012$V4016[incident2012$V4016 >= 997] <- NA
  
```



## Create a "date year" weight
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  i <- which(incident2012$V4019 == "(2) No (is series)")
  incident2012$WGTVICDY <- incident2012$WGTVICCY
  incident2012$WGTVICDY[i] <- with(incident2012, WGTVICDY[i] * V4016[i])
  
```



## Standardize coding of crime type
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  incident2012$V4529 <- gsub("\\(([1-9])))", "(0\\1)", incident2012$V4529)
  
```



# Create 2012 household dataframe
## Bind 2012 and 2013 incident data
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  household2012 <- rbind(househld12, househld13)
  household2012 <- subset(household2012, YEARQ >= 2012.1 & YEARQ <= 2013.2)

```



## Standardize coding of "month allocated" variable
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  household2012$V2130 <- gsub("\\(0", "\\(", household2012$V2130)
  table(household2012$V2130)

```



# Create 2012 person dataframe
## Fix incompatible factor and numeric values between 2012 and 2013
2012 factor levels look like "(1) Yes" but coded only as "1" in 2013.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  # gives factor levels for each variable
    i <- sapply(person12, levels)

  # gives factor levels or each factor variable, non-factor variables return null result
    i <- i[!sapply(i, is.null)]
  
  # store in i, variable where factor levels begin with "("
    i <- sapply(i, function(x) all(substring(x,1,1) == "("))
    var.fix <- names(i)[i]
    
  # for-loop to fix names
    for(xj in var.fix)
    {
      person12[,xj] <- gsub("\\(([0-9]+)\\).*", "\\1", person12[,xj]) # remove words that follow the parenthesis
      person12[,xj] <- as.numeric(person12[,xj])
    }

```



## Stack 2012 and 2013 data 
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  person2012 <- rbind(person12, person13)
  person2012 <- subset(person2012, YEARQ >= 2012.1 & YEARQ <= 2013.2)

```



# Merge incident and person dataframes
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  # Match and merge data
    a <- merge(incident2012,                      # indicent data
               person2012[, c("IDPER", "YEARQ",   # unique ID's of person
                            "V3014",              # age
                            "V3015",              # marital status
                            "V3018")],            # sex
               by = c("IDPER", "YEARQ"),          # variables to match/merge
               all.x = TRUE)                      # keep all incidents even if not matched
  
    incident2012 <- a

  # Rename age, marital, status, and sex variables
    names(incident2012)[names(incident2012) == "V3014"] <- "age"
    names(incident2012)[names(incident2012) == "V3015"] <- "marital"
    names(incident2012)[names(incident2012) == "V3018"] <- "sex"
    
  # View first line of incident data to see if the merge worked!
    incident2012[1, c("IDPER", "YEARQ", "age", "marital", "sex")]

```


## Create new variable: age categories
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  incident2012$ageGroup <- cut(incident2012$age, breaks = c(0,16,21,35,45,60,110))

```


## Assign number values to actual names to age and marital status variables
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  # Deal with missing value indicators
    incident2012$marital[incident2012$marital == 8] <- NA

  # Add labels
    incident2012$marital <- factor(incident2012$marital, levels = 1:5, 
                                   labels = c("married", "widowed", "divorced", "separated", "never married"))
    
    incident2012$sex <- factor(incident2012$sex, levels = 1:2,
                               labels = c("male", "female"))

```



# Reshape data to find out common crime types by sex (of the crime victim)
## Estimated counts (weighted)
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  a <- aggregate(WGTVICDY ~ V4529 + sex, data = incident2012, FUN = sum)
  a <- reshape(a, timevar = "sex", idvar = "V4529", direction = "wide")
  a[is.na(a)] <- 0
  names(a) <- c("crimeType", "male", "female")
  a

```



## Compute row percentages to determine what percentage of each crime is male and female
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

  temp <- a
  row.total <- with(temp, male + female)
  temp$male <- with(temp, 100 * male / row.total)
  temp$female <- with(temp, 100 * female / row.total)
  rowSums(temp[,-1])  # check that the rows sum to 100
  
  temp$ratio <- temp$female / temp$male
  temp[order(-temp$ratio),]

```