---
title: "Working with SQL in R: Chicago Crime Incident Data (2001-present)"
author: "Pauline I. Alvarado"
date: "Fall 2018"
output: pdf_document
---

This coding exercise was from [University of Pennsylvania's Criminal Justice Data Science Course](https://github.com/gregridgeway/R4crim>) taught by Dr. Gregory Ridgeway. Script has been modified from course examples to fit personal notation conventions. Data was taken from the [Chicago Data Portal](https://data.cityofchicago.org/Public-Safety/Crimes-2001-to-present/ijzp-q8t2).  

# Clean Data

## Load CSV File and Check Commas
Values should be separated by semi-colon instead of commas. Extra commas in SQL can create a new column affecting the shape of the dataset.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  # Read first 1000 lines
    infile  <- file("chicago-crime-2001-pres.csv", "r") # set up ability to read file
    a <- readLines(infile, n=10000)
    close(infile) # close the file

  # Count commas for each row and check consistency
    a[1]
    regexpr(",",a[1])
    gregexpr(",",a[1])

    b <- sapply(gregexpr(",",a),length)
    table(b)

    i <- which(b == 21)
    a[2] # 22 commas
    a[i[1:4]]
    
    i <- which(b == 23)
    a[2] # 22 commas
    a[i[1:4]]
    
    i <- which(b == 24)
    a[2] # 22 commas
    a[i[1:4]]
    
```


## Use regular expressions to clean data
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  # Set up input file and output file
    infile  <- file("chicago-crime-2001-pres.csv",       'r')
    outfile <- file("chicago-crime-2001-pres-clean.csv", 'w')
    
  # Fix variable names
    a <- readLines(infile, n = 1)
    a <- gsub(",Location$", "", a) # just delete ",Location" at end
    a <- gsub(",", ";", a) # separate with ;
    a <- gsub("[\\. ]", "", a)  # SQL doesn't like field names with .
    writeLines(a, con=outfile)
    flush(outfile) # flush what's been written to the storage media

  # Loop through the rest of the data, 100,000 rows at a time
    c.lines <- 0
    system.time(
      while ((length(a <- readLines(infile, n=100000)) >  0))
      {
        c.lines <- c.lines + length(a)
        print(c.lines)
        
        # Remove the redundant "Location" coordinates
          a0 <- gsub(",\"\\([[:print:]]*","", a) # Delete everything after parenthesis
  
        # Remove the extra comma for those missing Location
         i <- which(a == a0)
         a0[i] <- gsub(",$", "", a0[i])
         a <- a0
  
        # Turn commas outside double quotes to semicolons
        # Use ?= to "lookahead" for paired quotes
          a <- gsub("(,)(?=(?:[^\"]|\"[^\"]*\")*$)", ";", a, perl = TRUE)

        # Write the cleaned up data to storage
          writeLines(a, con=outfile)
      }
    )
    close(infile)
    close(outfile)

  # Check any problematic lines
    a <- scan("chicago-crime-2001-pres-clean.csv", what="",sep="\n",skip=850000,nlines=3)
    
```



# Build a SQL Database
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Load Package
    library(sqldf)

  # Import the cleaned data file into RSQLite
    con <- dbConnect(SQLite(), dbname="chicagocrime.db")
    a <- read.table("chicago-crime-2001-pres-clean.csv", sep = ";", nrows = 5, header = TRUE)
    variabletypes <- dbDataType(con, a)
    variabletypes["IUCR"] <- "TEXT"
    
    if(dbExistsTable(con, "crime")) dbRemoveTable(con, "crime")
  
    dbWriteTable(con, "crime",
                 "chicago-crime-2001-pres-clean.csv",
                 row.names = FALSE,
                 header = TRUE,
                 field.types = variabletypes,
                 sep = ";") 
    
    # Check if the new "crime" table exists and view columns
      dbListFields(con, "crime") 
    
    # Disconnect to finalize
      dbDisconnect(con) 
    
```



## Clean data via SQL query
### Missing Values
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Convert blanks into NULL
    con <- dbConnect(SQLite(), dbname="chicagocrime.db")
    res <- dbSendQuery(con, "UPDATE crime 
                             SET Ward = NULL 
                             WHERE (Ward = '')")
    res <- dbSendQuery(con, "UPDATE crime 
                             SET District = NULL 
                             WHERE (District = '')")
    
  # Convert empty coordinate values into NULL
    res <- dbSendQuery(con, "UPDATE crime 
                             SET XCoordinate = NULL
                             WHERE (YCoordinate = '') OR (YCoordinate = 0)")
    res <- dbSendQuery(con, "UPDATE crime 
                             SET YCoordinate = NULL
                             WHERE (YCoordinate = '') OR (YCoordinate = 0)")
    
  # Round up figures close to 0
    res <- dbSendQuery(con, "UPDATE crime
                             SET Latitude = NULL
                             WHERE (Latitude = '') OR (ABS(Latitude-0.0)<0.01)")
    res <- dbSendQuery(con, "UPDATE crime
                             SET Longitude = NULL
                             WHERE (Longitude = '') OR (ABS(Longitude-0.0)<0.01)")
    

```


### Remove coordinates outside of Chicago
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  res <- dbSendQuery(con, "UPDATE crime 
                           SET Longitude = NULL
                           WHERE Longitude < -88")
  res <- dbSendQuery(con, "UPDATE crime 
                           SET Latitude = NULL
                           WHERE Latitude < 41")
  
  dbDisconnect(con) 

```



# Create IUCR Lookup Table
Primary crime type, FBI code, and IUCR are redundant in one table. Put in separate tables but link all of the information.
## Remove redundant information
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Load Packages
    library(gsubfn)
    library(proto)
    library(RSQLite)

  # Eliminate duplicate rows
    con <- dbConnect(SQLite(), dbname="chicagocrime.db")
    res <- dbSendQuery(con, "SELECT DISTINCT IUCR, PrimaryType, FBICode
                             FROM crime")
    a <- fetch(res, n = -1)
    dbClearResult(res)
    
  # Check that each IUCR is linked to a unique value primary crime type and FBI code
    sort(table(a$IUCR))
    subset(a, IUCR %in% 2091:2093)
    
```



### Research discrepancy with FBI codes
#### Narcotics
IUCR codes 2091 (forfeit property), 2092 ("soliciting narcotics on a publicway)" and 2093 ("found suspect narctics") are related to narcotics. Accompanying FBI codes include 18 ("crimes of production, sale, use of drugs) and 26 ("miscellaneous"). Usage of FBI Code 26 phases out after 2015.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  res <- dbSendQuery(con, "SELECT DISTINCT IUCR,
                                         FBICode,
                                         SUBSTR(Date,7,4) AS year
                         FROM crime
                         WHERE IUCR='2091' OR IUCR='2092' OR IUCR='2093'
                         ORDER BY IUCR,FBICode, year")
  fetch(res, n = -1)
  dbClearResult(res)
  
```



#### Offenses involving children
IUCR crimes 1710, 1715, 1725, 1755, and 1780 are offenses that involve children. Before 2016, assigned as FBI Code 26 and recently coded as FBI Code 20 ("offense against family").
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  res <- dbSendQuery(con, "SELECT COUNT(*) as crimecount,
                                  IUCR,
                                  FBICode,
                                  PrimaryType,
                                  SUBSTR(Date,7,4) AS year
                         FROM crime
                         WHERE IUCR='1710' OR IUCR='1715' OR IUCR='1725' OR IUCR='1755' OR IUCR='1780'
                         GROUP BY IUCR, PrimaryType, FBICode, year")
  fetch(res, n = -1)
  dbClearResult(res)
  
```


#### Possession of incendiary devices
IUCR crimes 1030 and 1035 are offenses that involve possession of incendiary devices. Recently coded as arson instead of miscellaneous.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  res <- dbSendQuery(con, "SELECT DISTINCT IUCR,
                                         FBICode,
                                         SUBSTR(Date,7,4) AS year
                         FROM crime
                         WHERE IUCR='1030' OR IUCR='1035'
                         ORDER BY IUCR,FBICode, year")
  fetch(res, n = -1)
  dbClearResult(res)
  
```




#### "NON-CRIMINAL" values within PrimaryType
Extra spaces were removed in the phrase.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  res <- dbSendQuery(con, "SELECT COUNT(*) as crimecount,
                                  IUCR,
                                  FBICode,
                                  PrimaryType,
                                  SUBSTR(Date,7,4) AS year
                         FROM crime
                         WHERE IUCR='5114'
                         GROUP BY IUCR, PrimaryType, FBICode, year")
  fetch(res, n = -1)
  dbClearResult(res)
  
```



## Create IUCR, PrimaryType, FBICode lookup table within Chicago crime database
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  if(dbExistsTable(con, "iucr")) dbRemoveTable(con, "iucr")

  # import dataframe into SQLite
    dbWriteTable(con, "iucr", a, row.names = FALSE)
    dbListFields(con, "iucr")
    
  # check if the table is correct
    fetch(dbSendQuery(con, "SELECT * from iucr LIMIT 5"))
    
  # clear rows with redudant and outdated FBI codes and removal of spaces for IUCR 5114
    res <- dbSendQuery(con, "DELETE from iucr
                             WHERE (FBICode = '26') AND
                                    (iucr IN ('1030', '1035',
                                             '1710', '1715', '1725', '1780',
                                             '2091', '2092', '2093'))")
    res <- dbSendQuery(con, "DELETE from iucr
                             WHERE  (iucr = '5114') AND
                                    (PrimaryType = 'NON - CRIMINAL')")
    
  # check that database has two tables
    dbListTables(con)
    
```



## Add dates in standard format to the database
### View date-related values
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  library(lubridate)

  # View current format
    res <- dbSendQuery(con, "SELECT ID, Date FROM crime")
    data <- fetch(res, n = -1)
    dbClearResult(res)
    data[1:5,]
  
  # View variable classification
    res <- dbSendQuery(con, "PRAGMA table_info(crime)")
    fetch(res, n = -1)
    dbClearResult(res)
    
```



### Transform dates in standard format
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  data$datefix <- mdy_hms(data$Date)
  data$datefix <- as.character(data$datefix)
  data[1:5,]
  data$Date <- NULL #delete original date from the dataframe
  
  # Create new date table
     if(dbExistsTable(con, "DateFix")) dbRemoveTable(con, "DateFix")
    dbWriteTable(con, "DateFix", data, row.names = FALSE)
    dbListTables(con)
    
   
```




# Finalize the crime table with complete information
## Create a new crime table
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
    
  # Rename "crime" table to "crime_old" (will be deleted)
    if(dbExistsTable(con, "crime_old")) dbRemoveTable(con, "crime_old")
    res <- dbSendQuery(con, "ALTER TABLE crime RENAME TO crime_old")
    dbClearResult(res)
    
  # Create new "crime" table
    a <- dbListFields(con, "crime_old")
    paste(a, collapse = ",")
    
    res <- dbSendQuery(con, "CREATE TABLE crime AS
                             SELECT crime_old.ID,
                                    crime_old.CaseNumber,
                                    DateFix.datefix AS date,
                                    crime_old.Block,
                                    crime_old.IUCR,
                                    crime_old.Description,
                                    crime_old.LocationDescription,
                                    crime_old.Arrest,
                                    crime_old.Domestic,
                                    crime_old.Beat,
                                    crime_old.District,
                                    crime_old.Ward,
                                    crime_old.CommunityArea,
                                    crime_old.XCoordinate,
                                    crime_old.YCoordinate,
                                    crime_old.Latitude,
                                    crime_old.Longitude
                              FROM  crime_old, DateFix
                              WHERE crime_old.ID=DateFix.ID") # link two tables together
    dbClearResult(res)
    
  # View new table
    res <- dbSendQuery(con, "SELECT *
                             FROM crime")
    fetch(res, n = 10)
    dbClearResult(res)
    
  # Delete crime_old and DateFix
    res <- dbSendQuery(con, "DROP TABLE crime_old")
    dbClearResult(res)
    res <- dbSendQuery(con, "DROP TABLE DateFix")
    dbListTables(con)
    
  # Clean up unused space
    system.time(res <- dbSendQuery(con, "VACUUM"))
    dbClearResult(res)

```



# Join data across tables
## Create an "index"
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  timeIUCRjoin <- system.time(
    {
      res <- dbSendQuery(con, "SELECT crime.CaseNumber,
                                      iucr.FBIcode
                               FROM crime,
                                    iucr
                               WHERE crime.iucr = iucr.iucr")
      data <- fetch(res, n = -1)
    })
  dbClearResult(res)
  data[1:10,]

```


## Use INNER JOIN
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Tabulate number of Part 1 crimes occuring in each year
    res <- dbSendQuery(con, "SELECT iucr.PrimaryType AS     type,
                                    STRFTIME('%Y', date) AS year,
                                    COUNT(*) AS             crimecount
                             FROM crime
                                  INNER JOIN iucr
                                  ON crime.iucr = iucr.iucr
                              WHERE FBICode IN ('01A', '02', '03', '04A', '04B', 
                                                '05', '06', '07', '09')
                              GROUP BY type, year")
    fetch(res, n = -1)
    dbClearResult(res)

```



# SUBQUERIES
Results from one query as part of another query
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Determine whether a block shows up multiple times in initial query
    res <- dbSendQuery(con, "SELECT COUNT(*), Block
                             FROM
                                (SELECT DISTINCT Block,
                                                 XCoordinate,
                                                 YCoordinate
                                 FROM crime)
                             GROUP BY Block")
    fetch(res, n = 20)
    dbClearResult(res)
    
```




# Exercises
## Which district has the most thefts?
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  res <-dbSendQuery(con, "SELECT    COUNT(*) as crimecount,
                                    District
                          FROM      crime
                                    INNER JOIN iucr ON iucr.iucr
                          WHERE     iucr.PrimaryType = 'THEFT'
                          GROUP BY  District")
  a <- fetch(res, n = -1)
  dbClearResult(res)
  
  a[which.max(a$crimecount), ]

```



## Count number of assaults, since 2010, that occured on Fridays and Saturdays, after 6pm
Report date, day of week, hou of the day, and year.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

    res <- dbSendQuery(con, "SELECT   COUNT(*),
                                      DATE(crime.Date) AS crimedate,
                                      CAST(STRFTIME('%w', crime.Date) AS INTEGER) AS weekday,
                                      CAST(STRFTIME('%H', crime.Date) AS INTEGER) AS hour,
                                      CAST(STRFTIME('%Y', crime.Date) AS INTEGER) AS year
                            FROM      crime
                                          INNER JOIN iucr ON crime.iucr = iucr.iucr
                            WHERE     iucr.PrimaryType = 'ASSAULT' AND
                                      year >= 2010 AND
                                      (weekday >= 5) AND (weekday <= 6) AND
                                      hour >= 18
                            GROUP BY  crimedate, weekday, hour, year")
  fetch(res, n = -1)
  dbClearResult(res)
                           
                     
```

