---
title: "San Francisco Crime Density Map (2003-2018)"
author: "Pauline I. Alvarado"
date: "12/19/2018"
output: pdf_document
---

This coding exercise was submitted for the [University of Pennsylvania's Criminal Justice Data Science Course](https://github.com/gregridgeway/R4crim>) taught by Dr. Gregory Ridgeway. Code has been modified from course examples to fit the needs of dataset and description of process. Data consists of [San Francisco Police Department's Incident Reports from 2003 to May 2018](https://data.sfgov.org/Public-Safety/Police-Department-Incident-Reports-Historical-2003/tmnf-yvry) taken from the city of San Francisco's Open Data site. There are 2.21 million incidents included in this .csv file.  

Key skills: Data Preprocessing, Data Wrangling, Creating SQLite Databases, SQL Queries, Regular Expressions, Density/Hot Spot Mapping


# Data Preprocessing and Wrangling
### Load packages and view the first 5 rows of the data set.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

    library(sqldf)
    library(ggmap)
    scan("sfpd-incident-2003to2018.csv", what="", nlines=5, sep="\n")

```



### Set up a "while loop" to ensure that commas and semicolons are correctly used in addition to removing periods in field names.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  # Create input and output files
    infile  <- file("sfpd-incident-2003to2018.csv",       'r')
    outfile <- file("sfpd-incident-2003to2018-clean.csv", 'w')
      
  # Get variable names & make the symbols SQL friendly
    a <- readLines(infile, n = 1)
    a <- gsub(",", ";", a)
  
    writeLines(a, con = outfile)
    flush(outfile)
      
  # Loop through the rest of the data, 100,000 rows at a time
    c.lines <- 0
  
    system.time(
        while ((length(a <- readLines(infile, n=100000)) >  0))
        {
            c.lines <- c.lines + length(a)
            print(c.lines)
  
        # Turn commas outside double quotes to semicolons and use ?= to "lookahead" for paired quotes
        a <- gsub("(,)(?=(?:[^\"]|\"[^\"]*\")*$)",";", a, perl = TRUE)
        
        # Manually handle problematic lines
        a <- gsub("AEROSOL CONTAINER; SALE, PURCHASE OR POSSESSION OF",
                  "AEROSOL CONTAINER, SALE, PURCHASE OR POSSESSION OF", a)
        a <- gsub("DOG, FIGHTING;  OWNING, FIGHTING, OR ATTENDING FIGHT",
                  "DOG, FIGHTING, OWNING, FIGHTING, OR ATTENDING FIGHT", a)
          
        writeLines(a, con=outfile)
        }
    )
    
    close(infile)
    close(outfile)

```



# Set Up SQL Database
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  con <- dbConnect(SQLite(), dbname="sfcrime.db")
  a <- read.table("sfpd-incident-2003to2018-clean.csv", sep = ";", nrows = 5, header = TRUE)
  variabletypes <- dbDataType(con, a)
    
    
```


## Import the cleaned data file into RSQLite
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  if(dbExistsTable(con, "crime")) dbRemoveTable(con, "crime")

  dbWriteTable(con, "crime",
               "sfpd-incident-2003to2018-clean.csv",
               row.names = FALSE,
               header = TRUE,
               field.types = variabletypes,
               sep = ";") 
    
```


### Check if the new "crime" table exists and view columns
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  dbListFields(con, "crime")
    
```


### Disconnect to finalize
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  dbDisconnect(con)
    
```



# Create a Crime Hot Spot / Density Map
### Connect to the SQL database and examine min/max coordinates in the dataset
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  con <- dbConnect(SQLite(), dbname="sfcrime.db")
      
  res <- dbSendQuery(con, "
                         SELECT MIN(X)  AS min_lat,
                                MAX(X)  AS max_lat,
                                MIN(Y) AS min_lon,
                                MAX(Y) AS max_lon
                          FROM  crime
                          WHERE (X IS NOT NULL) AND
                                (Y IS NOT NULL)")
  fetch(res, n = -1)
  dbClearResult(res)

```


### Remove data outside of city limits
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
res <- dbSendQuery(con, "
                   UPDATE crime SET Y=NULL
                   WHERE Y > 37.9")
dbClearResult(res)

```


### Create a SQL Query to get data for mapping
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  res <- dbSendQuery(con, "
                     SELECT X,Y
                     FROM crime")
      a <- fetch(res, n = -1)
      dbClearResult(res)
      
```


### Get a map of San Francisco taken from Stamen and add some random points to the map
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  sf.map <- ggmap(get_map(c(-122.52, 37.7, -122.35, 37.81), scale = "auto",source = "stamen"))
  
  i <- sample(1:nrow(a),250000)
      sf.map + 
        geom_point(aes(x = X,y = Y), data = a[i,], alpha = 0.5, color = "darkred", size = 0.1)
    
```


### Format a hot spot / density map
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  sf.map +
    stat_density2d(aes(x = X, y = Y, fill =..level.., alpha = ..level..),
                   bins = 60, data = a[i,], geom = 'polygon') +
    scale_fill_gradient('Crime Density',low = "green",high = "red") +
    scale_alpha(range = c(.4, .75), guide = FALSE) +
    guides(fill = guide_colorbar(barwidth = 1.5, barheight = 10)) +
    labs(title = "San Francisco Crime Hotspot Map (2003 - 2018)",
         subtitle = "Number of Crime Incidents Reported to San Francisco Police Department",
         x = "Longitude",
         y = "Latitude") +

    theme_bw() +
    
    theme(panel.border = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),

          axis.line = element_line(color = "#D8D8D8"),
          axis.ticks.y = element_blank(),
          axis.ticks.x = element_blank(),

          plot.title = element_text(color = "#404040", size = 12, face = "bold"),
          plot.subtitle = element_text(color = "#404040", size = 9, face = "italic"),
          axis.title.x = element_text(color = "#404040", size = 9, face = "bold"),
          axis.title.y = element_text(color = "#404040", size = 9, face = "bold"),
          
          strip.text.x = element_text(size=9, face = "bold"),
          strip.background = element_rect(colour="#D8D8D8", fill="#D8D8D8")
      )
    
```

