---
title: "Criminal Justice Analytics Final Project: 2017 Los Angeles Parking Citations"
author: "Pauline I. Alvarado"
date: "12/13/2018"
output: pdf_document
always_allow_html: true
---

This final project was submitted for [University of Pennsylvania's Criminal Justice Data Science Course](https://github.com/gregridgeway/R4crim>) taught by Dr. Gregory Ridgeway, who conducted the one-on-one code review. Data consists of parking citations from the city of Los Angeles in 2017 taken from the [LA Open Data Website](https://data.lacity.org/A-Well-Run-City/Parking-Citations/wjz9-h9np/data). Available data, to date (2010-2018), consists of more than 8 million entries. The filter function on the website was used to download only 2017 data. California city shapes, California Census tracts, and Los Angeles County Rods were acquired from the [Census](https://www.census.gov/cgi-bin/geo/shapefiles/index.php) website.


# Research Question
In which areas in Los Angeles is a person most likely to get a ticket from an expired parking meter? 


# Data Management
## Load Relevant Packages and Review Data
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Load package
    library(sqldf)

  # Review data
    scan("parking-citations-2017.csv", what = "", nlines = 10, sep = "\n")
    
```


## Clean Up the File
Space between words need to be moved for variables. Multiple commas indicate no data and need to be converted to NULL.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
   
  # Set up ability to read the file
    infile  <- file("parking-citations-2017.csv",       'r')
    outfile <- file("parking-citations-2017-clean.csv", 'w')

  # Clean up column names
    a <- readLines(infile, n=1) # Get variable names from Row 1
    a <- gsub(",",";",a) 
    a <- gsub(" ","",a)  
    writeLines(a, con=outfile)
    flush(outfile)

  # Clean the rest of the data with a loop 
    c.lines <- 0
  
    system.time(
      while ((length(a <- readLines(infile, n=100000)) >  0))
        {
          c.lines <- c.lines + length(a)
          print(c.lines)
          
          # Change all commas to a semi-colon, lookahead looks for paired quotes
          a <- gsub("(,)(?=(?:[^\"]|\"[^\"]*\")*$)",";",a,perl=TRUE)
          
          # Remove semi-colon for a specific line
          a <- gsub("605; WILSHIRE", "605 WILSHIRE", a)
  
          # Write the cleaned up data to storage
          writeLines(a, con=outfile)
        }
    )
    close(infile)
    close(outfile)
  
  # Spotcheck lines
    scan("parking-citations-2017-clean.csv", what="",sep="\n",skip=339410,nlines=3)

```   
    
    
## Build a SQL Database
### Set up the database
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
    
  # Look over the first few rows of the dataset
    scan(what="",file="parking-citations-2017-clean.csv",nlines=5,sep="\n")
    a <- read.table("parking-citations-2017-clean.csv",sep=";",nrows=5,header=TRUE)

  # Run once to set up a database
    # Connect or create a new GSQLite database
      con <- dbConnect(SQLite(), dbname="la-parking-citations-2017.db")
      variabletypes <- dbDataType(con, a)
      if(dbExistsTable(con, "parking")) dbRemoveTable(con, "parking")
      dbWriteTable(con, "parking",
                   "parking-citations-2017-clean.csv",
                   row.names=FALSE,
                   header=TRUE,
                   field.types=variabletypes,
                   sep=";") 
      dbListFields(con,"parking")
      dbDisconnect(con)

```


### Format Date/Time Column
#### Load related package and view column characteristics
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Related package
    library(lubridate)
      
  # Info on internal data
    con <- dbConnect(SQLite(), dbname="la-parking-citations-2017.db") # reconnect to database
    res <- dbSendQuery(con, "PRAGMA table_info(parking)")
    fetch(res, n = -1)
    dbClearResult(res)

      
```


#### Convert to proper date format
Date column is currently categorized as "TEXT".
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Convert to proper date format
    res <- dbSendQuery(con, "SELECT Ticketnumber, IssueDate FROM parking")
    data <- fetch(res, n = -1)
    dbClearResult(res)

    data$datefix <- mdy(data$IssueDate)
    data$datefix <- as.character(data$datefix)
    data$Date <- NULL

    if(dbExistsTable(con,"DateFix")) dbRemoveTable(con, "DateFix")
    dbWriteTable(con, "DateFix", data, row.names=FALSE)
    dbListTables(con)

  # Inspect the new DateFix table
    res <- dbSendQuery(con, "
                            SELECT TicketNumber, IssueDate,
                            datefix,
                            strftime('%w',datefix)
                            FROM DateFix")
    fetch(res, n = 10)
    dbClearResult(res)
      
  # Check if '%w' work
    wday("2017-01-01",label=TRUE)

  # Drop the old parking table & create a new parking table with formatted dates
      # Rename old table
      res <- dbSendQuery(con, "
                              ALTER TABLE parking RENAME TO parking_old")
                              dbClearResult(res)
      dbListTables(con) # view all existing tables

      # List all variables to keep
        a <- dbListFields(con,"parking_old")
        paste(a,collapse=",")
  
        res <- dbSendQuery(con, "
                                CREATE TABLE parking AS
                                SELECT parking_old.Ticketnumber,
                                        DateFix.datefix AS IssueDate,
                                        parking_old.Issuetime,
                                        parking_old.MeterId,
                                        parking_old.MarkedTime,
                                        parking_old.PlateExpiryDate,
                                        parking_old.VIN,
                                        parking_old.Make,
                                        parking_old.BodyStyle,
                                        parking_old.Color,
                                        parking_old.Location,
                                        parking_old.Route,
                                        parking_old.Agency,
                                        parking_old.Violationcode,
                                        parking_old.ViolationDescription,
                                        parking_old.Fineamount,
                                        parking_old.Latitude,
                                        parking_old.Longitude
                                FROM parking_old,DateFix
                                WHERE parking_old.Ticketnumber=DateFix.Ticketnumber")
        
      # Look at the new table
        res <- dbSendQuery(con, "
                                SELECT *
                                FROM parking")
        fetch(res, n = 10)
        dbClearResult(res)
      
      # Look at all the existing tables & drop unecessary tables
        dbListTables(con)
        
        res <- dbSendQuery(con, "DROP TABLE parking_old")
        dbClearResult(res)
  
        res <- dbSendQuery(con, "DROP TABLE DateFix")
        dbClearResult(res)

  # Clean up unused space
    system.time(res <- dbSendQuery(con, "VACUUM"))

```

 
# Select Data to Analyze
## View top 50 violations
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
    
  # Reconnect to the database
    con <- dbConnect(SQLite(), dbname="la-parking-citations-2017.db")

  # Test query
    res <- dbSendQuery(con, "SELECT COUNT (*), ViolationDescription
                            FROM parking
                            GROUP BY ViolationDescription
                            ORDER BY COUNT(*) DESC
                            LIMIT 50")
    top50_violations <- fetch(res, n = -1)
    dbClearResult(res)
    dbDisconnect(con)

    top50_violations
    
```


## Create working dataset
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
      
  con <- dbConnect(SQLite(), dbname="la-parking-citations-2017.db")
  res <- dbSendQuery(con, "SELECT Latitude,Longitude, ViolationDescription
                            FROM parking
                            WHERE 
                                  ViolationDescription='METER EXP.' OR
                                  ViolationDescription='METER EXPIRED'")
  meter_expired_2017 <- fetch(res, n = -1)
  dbClearResult(res)
  dbDisconnect(con)

  
  # Remove Latitude and Longitude null value of 99999
    meter_expired_2017 <- subset(meter_expired_2017, Latitude!="99999")
    

```



# Plot Expired Meter Citation Data
## Ensure both the expired meter data and shapefile are on the coordinate system

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Load package to work with simple features files
    library(sf)
      
  # Load map of Los Angeles shapefile
    map_ca <- st_read("tl_2014_06_place/tl_2014_06_place.shp")
    map_la <- subset(map_ca, NAMELSAD=="Los Angeles city")
    
  # Add simple features data to meter_expired_2017 dataframe
    meter_expired_shape <- st_as_sf(meter_expired_2017,
                     coords=c("Latitude","Longitude"),
                     crs="+proj=lcc +lat_1=34.03333333333333 +lat_2=35.46666666666667 
                          +lat_0=33.5 +lon_0=-118 +x_0=2000000 +y_0=500000.0000000002 
                          +ellps=GRS80 +datum=NAD83 +to_meter=0.3048006096012192 no_defs") 
      meter_expired_shape<- st_transform(meter_expired_shape, crs = 4269)
    
  # View coordinate systems
    st_crs(map_la)
    st_crs(meter_expired_shape)
    
    
```


## Plot simple features map
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

    plot(st_geometry(map_la))
    plot(st_geometry(meter_expired_shape), pch=".", col="green", add=TRUE)
    
```


## Plot point data on dynamic map using Leaflet
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Load Leaflet
    library(leaflet) 

 # Extract coordinates
    meter_expired_leaf <- st_coordinates(meter_expired_shape)
    meter_expired_leaf <- data.frame(meter_expired_leaf)
    names(meter_expired_leaf) <- c("longitude","latitude")
    
 # Aggregate by coordinates
   library(tidyverse)
   meter_expired_leaf <- meter_expired_leaf %>% group_by(latitude, longitude) %>% summarise(count = n())
   
  # Plot data on in a bubble map format
    leaflet(meter_expired_leaf) %>% 
    addTiles() %>%
    addCircles(lng = ~longitude, lat = ~latitude, weight = 1, radius = ~sqrt(count) * 10)
      
```
