---
title: "Working with FBI Uniform Crime Report Data: Offenses Known and Clearances by Arrest (2012)"
author: "Pauline I. Alvarado"
date: "Fall 2018"
output: pdf_document
---

This coding exercise was from [University of Pennsylvania's Criminal Justice Data Science Course](https://github.com/gregridgeway/R4crim>) taught by Dr. Gregory Ridgeway. Script has been modified from course examples to fit personal notation conventions. FData was taken from the [National Archive of Criminal Justice Data (NACJD)](https://doi.org/10.3886/ICPSR35021.v1).  


# Data Wrangling

## Load and view data

### Overview
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  # Load data
    load("UCR2012/DS0001/35021-0001-Data.rda")

  # View data/object components
    is(da35021.0001)
    
  # View first 2 line and 200 columns
    da35021.0001[1:2,1:200]

```



### View variable definitions stored in the data
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  var.lookup <- attributes(da35021.0001)$variable.labels
  var.lookup[1:32]

```



## Subset agencies that have complete data

### View which months have been reported
Result: Most of the data is reported in December
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  table(da35021.0001$V12)

```



### View agencies that don't report in December
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  # View agencies that don't report in December with less than 1 million in population
    with(subset(da35021.0001, V12 != "(12) Dec last reported"), summary(V14, digits = 10))

  # View agencies reporting information with jurisdiction of greater than 50k population
    subset(da35021.0001, (V12 != "(12) Dec last reported") & (V14 > 50000))[,c("V27", "V29", "V14", "V13", "V12")]
    
  # View whether data wa also reported in another month
    var.names <- grep("MONTH INCLUDED IN", var.lookup, value = TRUE)
    var.names <- names(var.names)
    for(xj in var.names) 
    print(table(da35021.0001[,xj]))

```



### Subset agencies with calendar year end reports (most complete data) 
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  da35021.0001 <- subset(da35021.0001, V12 == "(12) Dec last reported")

```



## Create a dataframe
### Keep only relevant variables
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  ucr <- data.frame(ORI    = as.character(da35021.0001$V3),
                    AGENCY = as.character(da35021.0001$V29),
                    AREA   = as.character(da35021.0001$V26),
                    POP    = da35021.0001$V14,
                    MONTHS = as.character(da35021.0001$V12),
                    STATE  = as.character(da35021.0001$V2),
                    stringsAsFactors = FALSE)
                    
```



### Create new variables for each Part I crime offense
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  # Murder (Counts)
    var.names <- grep("ACT NUM MURDER", var.lookup, value = TRUE)
    var.names <- names(var.names)
    ucr$murder <- rowSums(da35021.0001[,var.names])
    
  # Rape (Counts)
    var.names <- grep("ACT NUM RAPE", var.lookup, value = TRUE)
    var.names <- names(var.names)
    ucr$rape <- rowSums(da35021.0001[,var.names])
    
  # Robbery (Counts)
    var.names <- grep("ACT NUM ROBBRY", var.lookup, value = TRUE)
    var.names <- names(var.names)
    ucr$robbery <- rowSums(da35021.0001[,var.names])
    
  # Aggravated Assault (Counts)
    var.names <- grep("ACT NUM ASSLT", var.lookup, value = TRUE)
    var.names <- names(var.names)
    ucr$assault <- rowSums(da35021.0001[,var.names])
    
    var.names <- grep("ACT # SIMPLE ASSLT", var.lookup, value = TRUE)
    var.names <- names(var.names)
    ucr$assault <- ucr$assault - rowSums(da35021.0001[,var.names])
    
   # Burglary (Counts)
    var.names <- grep("ACT # BURGLARY", var.lookup, value = TRUE)
    var.names <- names(var.names)
    ucr$burglary <- rowSums(da35021.0001[,var.names])
    
  # Larceny (Counts)
    var.names <- grep("ACT # LARCENY", var.lookup, value = TRUE)
    var.names <- names(var.names)
    ucr$larceny <- rowSums(da35021.0001[,var.names])
    
  # Grand Theft Auto (Counts)
    var.names <- grep("ACT # VHC THEFT", var.lookup, value = TRUE)
    var.names <- names(var.names)
    ucr$gta <- rowSums(da35021.0001[,var.names])
                    
```



# Exercises
## Make a data frame of "New Jersey"
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  nj <- ucr[grep("New Jersey", ucr$STATE),]

```



## How many murders in New Jersey? How many robberies?
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  sum(nj$murder)
  sum(nj$robbery)
                    
```



## Find burlary rate per person in New Jersey
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  sum(nj$burglary) / sum(nj$POP)
                    
```



## Find burglary rate per 100,000
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  sum(nj$burglary) / sum(nj$POP) * 100000
                    
```



## Add new column to New Jersey dataset for burglary rate for each agency and sory by descending order
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  # Create burglary rate variable
    nj$burglaryrate <- nj$burglary / nj$POP * 100000 

  # Remove missing / error values
    nj <- subset(nj, !is.nan(burglaryrate) & !is.infinite(burglaryrate))
    
  # View data in descending order
    nj[with(nj, order(-burglaryrate)),]
                    
```



