---
title: "Extracting Data from Text and Geocoding: Office-involved Shootings (OIS) in Philadelphia"
author: "Pauline I. Alvarado"
date: " Fall 2018"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```


This coding exercise was done in [University of Pennsylvania's Criminal Justice Data Science Course](https://github.com/gregridgeway/R4crim>) taught by Dr. Gregory Ridgeway. Code has been modified from course examples to fit personal learning style. Data consists of [PPD Service Area Shapefiles](https://www.opendataphilly.org/dataset/police-service-areas) and information scraped from the [PPD Officer-involved Shooting Website](http://phillypolice.com/ois). 

Key skills: Web and document scraping, regular expressions, OpenStreetMap and ArcGIS geocoders, JSON, simple features, stack/match/merge data


# Scraping OIS Data from PPD Webpage
## Load packages and grab raw html 
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Packages
    library(lubridate)
    library(tm)
    library(pdftools)
    library(jsonlite)
    library(ggmap)
    library(sf)

  # Grab raw html
    ppd_website <- scan("http://phillypolice.com/ois", what = "", sep = "\n")
    ppd_website <- gsub("\t", "", ppd_website)
    head(ppd_website)
    
```


## Create dataframe
### Grab 2013-2018 data
Make sure to view a sample row to determine how to clean up the data. Each table row starts with <tr-id="yyyy-####"]. The next row is the URL that contains PDF with detailed information. Date and address are in the 3rd and 4th row. respectively.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  a <- ppd_website
  i <- grep("id=\"2018-2954", a)
  a[i + 0:9]
  
```



### Extract, clean, and add 2013-2018 data to ois dataframe
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  i <- grep("id=\"201(3|4|5|6|7|8)", a)
  
  ois <- data.frame(id = gsub("<[^>]*>", "", a[i+1]),
                    date = gsub("<[^>]*>", "", a[i+2]),
                    location = gsub("<[^>]*>", "", a[i+3]),
                    url = gsub("<td><a href=\"(.*)\" class=.*", "\\1", a[i+1]),
                    stringsAsFactors = FALSE)
  
```



### Extract, clean, and add  2007-2012 data to OIS dataframe
Table for these years provide no incident date. Incident location is 2 elements after the <tr> tag.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  i <- grep("id=\"20(07|08|09|10|11|12)", a)
  
  temp <- data.frame(id = gsub("<[^>]*>", "", a[i+1]),
                    date = NA,
                    location = gsub("<[^>]*>", "", a[i+2]),
                    url = gsub("<td><a href=\"(.*)\" class=.*", "\\1", a[i+1]),
                    stringsAsFactors = FALSE)
  
```



### Stack "ois" and "temp" data
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  ois <- rbind(ois, temp) 
  ois$id <- gsub(" ", "", ois$id) # clear out extra spaces
  ois$url <- paste0("http://www.phillypolice.com", ois$url) # add full URL
  ois[1:5,]
  
```


### Consistent "id" variable format
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  # View rows that do not follow the ##-## convention
    grep("^[^-]*$", ois$id, value = TRUE)

  # Add hypens
    ois$id[ois$id == "1630"] <- "16-30"
    ois$id[ois$id == "1730"] <- "17-30"
    ois$id[ois$id == "1822"] <- "18-22"
    
```



#### View rows with missing id
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  subset(ois, id == "")
    
```


#### Willows Avenue OIS
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  # View rows above and below
    i <- grep("5100 block of Willows Avenue", ois$location)
    ois[(i-1):(i+1),]
    
  # Check if PDFs are available for 14-13, 14-14, 14-15
    a <- try(scan("http://www.phillypolice.com/assets/crime-maps-stats/officer-involved-shootings/14-13.pdf"))
    a <- try(scan("http://www.phillypolice.com/assets/crime-maps-stats/officer-involved-shootings/14-14.pdf"))
    a <- try(scan("http://www.phillypolice.com/assets/crime-maps-stats/officer-involved-shootings/14-15.pdf"))
    
  # Add 14-15 and URL to Willows Avenue OIS
    ois$id[1+which(ois$id=="14-12")] <- "14-15"
    ois$url[ois$id=="14-15"] <- "http://www.phillypolice.com/assets/crime-maps-stats/officer-involved-shootings/14-15.pdf"


```



#### View "B St and Ontario St" and "A St and Louden St"
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  # View rows above and below
    i <- which(ois$id == "")
    ois[c((i[1]-1):(i[1]+1), (i[2]-1):(i[2]+1)), 1:3]

  # After checking possible PDfs, add missing ids, remove smart quotes, and fix URL
    ois$id[1+which(ois$id=="10-60")] <- "10-65"
    ois$location[ois$id=="10-65"] <- "B and Ontario St"
    ois$url[ois$id=="10-65"] <- "http://www.phillypolice.com/assets/crime-maps-stats/officer-involved-shootings/2010/10-65%20B%20and%20Ontario%20St.pdf"

    ois$id[1+which(ois$id=="10-74")] <- "10-76"
    ois$location[ois$id=="10-76"] <- "A and Louden St"
    ois$url[ois$id=="10-76"] <- "http://www.phillypolice.com/assets/crime-maps-stats/officer-involved-shootings/2010/10-76%20A%20and%20Louden%20St.pdf"
    
```



### Clean up characters or extra spaces
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  # location column
    ois$location <- gsub("&quot;", "", ois$location)
    ois$location <- gsub("&amp;", "", ois$location)
    ois$location <- gsub("^ *| *$", "", ois$location)
    
  # html
    ois$url <- gsub("&amp;", "&", ois$url)
    ois$url <- gsub(" ","%20",ois$url) # need %20 to indicate space
    
```



### Reformat dates
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  ois$date <- mdy(ois$date)
    
```



## View working ois dataframe
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  head(ois)
    
```



# Extracting Data from PDF Files

## View and isolate date from pdf
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  # Raw text
    pdf_file_0701 <- "http://www.phillypolice.com/assets/crime-maps-stats/officer-involved-shootings/2007/07-01%204400%20N%2017%20ST.pdf"
    pdf_text_0701 <- pdf_text(pdf_file_0701)
    pdf_text_0701 
  
  # Separate lines
    a <- strsplit(pdf_text_0701, split = "\n")[[1]] # used only \n for mac can be a combo of \r\n
    a
    
  # Apply mdy() to a
    a <- mdy(a)
    
```



## For-loops to extract text from pdfs
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  ois$text <- NA

  for(i in 1:nrow(ois))
  {
    a <- pdf_text(ois$url[i])
    a <- paste(a, collapse = "\n") # collapse into one page
    ois$text[i] <- a
  }

    
```



## For-loops to extract dates and add to missing date fields 
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  for(i in which(is.na(ois$date)))
  {
    a <- strsplit(ois$text[i], split = "\n")[[1]]
    a <- sort(mdy(a))[1]
    
    ois$date[i] <- as.character(a)
  }

    
```



## Check date column
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Find any missing fields
    sum(is.na(ymd(ois$date)))

  # Reconcile id number and year (e.g. id starting with 07 only have dates in 2007)
    aggregate(year(date) ~ substr(id,1,2), data = ois, FUN = unique)
    
```



# Geocoding OIS Locations

## Automate the reformatting of URLs (

### Open Street Maps Function
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  geocode_osm <- function(address)
  {
    a <- gsub(" +", "\\%20", address)
    a <- paste0("http://nominatim.openstreetmap.org/search/", a, "?format=json&addressdetails=0&limit=1")
    return( fromJSON(a) )
  }
    
```



###ArcGIS Function
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  geocode_arcgis <- function(address)
  {
    a <- gsub(" +", "\\%20", address)
    a <- paste0("https://geocode.arcgis.com/arcgis/rest/services/World/GeocodeServer/findAddressCandidates?f=json&singleLine=",a,"&outFields=Match_addr,Addr_type")
    return( fromJSON(a) )
  }
    
```



## Clean up addresses

### View OISs with missing locations
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  i <- which(ois$location %in% c("", "withheld", "Withheld"))
  ois$text[i]
    
```



### Add or correct missing locations
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  ois$location[ois$id=="16-18"] <- "3200 block of Wellington Street"
  ois$location[ois$id=="10-06"] <- "Howard and Grange Street"
  ois$location[ois$id=="08-06"] <- "200 block of Clapier Street"
  ois$location[ois$id=="08-18"] <- "900 block of E. Slocum Street"
  ois$location[ois$id=="08-30"] <- "700 block of W. Rockland Street"
  ois$location[ois$id=="08-40"] <- "5400 Jefferson Street"
  ois$location[ois$id=="08-60"] <- "3000 Memphis Street"
  ois$location[ois$id=="08-70"] <- "1300 block of S. 29th Street"
  ois$location[ois$id=="08-74"] <- "5600 block of N. Mascher Street"

```



### Other type of address corrections
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Insert "and" between cross streets
    ois$location[ois$id=="07-19"] <- "51st and Arch" # needs "and"

  # Shooting that didn't involve PPD officers
    ois$text[ois$id=="17-08"] # shooting didn't involve PPD officers
    ois <- subset(ois, id != "17-08")
    
  # Assign address with "blocks" at the midpoint, e.g. change "5400 block of Erdick St" to "5450 Erdick St"
    a <- gsub("00 block( of)?", "50", a, ignore.case=TRUE)
    a <- gsub("unit bl(oc)?k( of)?", "50", a, ignore.case=TRUE)

  # Alleys
    a <- gsub("Rear Alley of |near ", "", a, ignore.case = TRUE)
    
  # Add city and state to improve geocoding accuracy
    a <- paste0(ois$location, ", Philadelphia, PA")

```



## Geocode locations

### Run addresses through OpenStreetMap geocoder
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  ois_gc <- vector("list", nrow(ois)) #list to store geocoding results
  for(i in 1:nrow(ois))
  {
    ois_gc[[i]] <- geocode_osm(a[i])
    if(length(ois_gc[[i]]) == 0)
    {
      cat("Could not geocode address #", i, ":", a[i], "/n")
    }
  }

```


### Run failed addresses through ArcGIS geocoder
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  for (i in which(sapply(ois_gc, length) == 0)) 
  {
    ois_gc[[i]] <- geocode_arcgis(a[i])
  }

```



### Reconcile formatting of OSM and ArcGIS results to extract longitude and latitude
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  # Function to extract longitude and latitude
    b <- lapply(ois_gc, function(x)
    {
      if(is.null(x$candidates)) #OSM
      {
        a <- data.frame(lon = as.numeric(x$lon),
                        lat = as.numeric(x$lat),
                        score = as.numeric(x$importance),
                        loctype = paste(x$class, x$type, sep = ":"),
                        method = "osm",
                        addressGeo = x$display_name,
                        stringsAsFactors = FALSE)
      } else # ArcGIS
      { 
        a <- data.frame(lon = x$candidates$location[1, "x"],
                        lat = x$candidates$locatio[1, "y"],
                        score = x$candidates$score[1],
                        loctype = x$candidates$attributes$Addr_type[1],
                        method = "arcgis",
                        addressGeo = x$candidates$attributes$Match_addr[1],
                        stringsAsFactors = FALSE)
      }
      return(a)
    })

  # Combine OSM and ArcGIS results
    ois_gc <- do.call(rbind, b)
    
  # Add a column containing the original address
    ois_gc <- cbind(ois_gc, addressOrig = a)
    
  # View dataframe
    head(ois_gc)

```



### Double-check the longitude and latitude

#### Check all coordinates and examine outliers
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  # View all longitude and latitude
    stem(ois_gc$lat)
    stem(ois_gc$lon)
    
  # View outliers
    subset(ois_gc, lon < -80) # Philadelphia, MS
    
```



#### Run outlier coordinates through ArcGIS geocoder
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  i <- which(ois_gc$lon < -80)
  b <- lapply(ois_gc$addressOrig[i], geocode_arcgis)
  
  b <- lapply(b, function(x)
  {
    data.frame(lon = x$candidates$location[1, "x"],
               lat = x$candidates$locatio[1, "y"],
               score = x$candidates$score[1],
               loctype = x$candidates$attributes$Addr_type[1],
               method = "arcgis",
               addressGeo = x$candidates$attributes$Match_addr[1],
               stringsAsFactors = FALSE)
  })
  
  b <- do.call(rbind, b)
  
  ois_gc[i, names(b)] <- b
  ois_gc[i,]
    
```



### Geocoding issues with location type
Location type should be specific locations (office, shop, building, station, house, PointAddress, StreetAddress, StreetInt). Highway, neighborhood, city, Locality, and StreetName are too general. 
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  sort(table(ois_gc$loctype))

    
```



### Map Officer Involved Shootings in Philadelphia
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  # Place all essential information in one data frame
    ois$lon <- ois_gc$lon
    ois$lat <- ois_gc$lat
    
  # Get map of Philly
    map_philly <- get_map(c(-75.288486, 39.868285, -74.950965, 40.138251), source = "stamen")
    ggmap(map_philly, extent = "normal") +
      geom_point(aes(x = lon, y = lat), data = ois,
                 color = "red",
                 alpha = 0.5) +
      theme(axis.title.x = element_blank(),
            axis.title.y = element_blank())
  
```



# Working with Shapefiles and Coordinate Systems

## Plot OIS Data on the Philly Police Service Area (PSA) Maps
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  # Load shapefile and check map projection
    map_ppd <- st_read("Boundaries_PSA/Boundaries_PSA.shp")
    st_crs(map_ppd)
    
  # Plot
    plot(st_geometry(map_ppd), axes = TRUE)
    points(lat~lon, data = ois, col = rgb(1,0,0,0.5), pch = 16)
  
```


## Exercise: Adding visual features to a map
### Add PSA number labels & highlight PSA areas
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Plot PSA map
    plot(st_geometry(map_ppd))
    axis(side =1) # x-axis
    axis(side = 2) # y-axis
  
  # Add number labels (center aligned within the polygon)
    a <- st_coordinates(st_centroid(st_geometry(map_ppd))) # extract coordinates from polygon
    text(a[,1], a[,2], map_ppd$PSA_NUM, cex = 0.5)
    
    
  # Highlight PSA 77 & 183
    text(a[,1], a[,2], map_ppd$PSA_NUM, cex = 0.5)
    plot(st_geometry(subset(map_ppd, PSA_NUM == "077")),
         add = TRUE, border = "red")
    plot(st_geometry(subset(map_ppd, PSA_NUM == "183")),
         add = TRUE, border = "green")
    
```



### Add PSA number labels & highlight PSA areas
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Plot PSA map
    plot(st_geometry(map_ppd))
    axis(side =1) # x-axis
    axis(side = 2) # y-axis
  
  # Add number labels (center aligned within the polygon)
    a <- st_coordinates(st_centroid(st_geometry(map_ppd))) # extract coordinates from polygon
    text(a[,1], a[,2], map_ppd$PSA_NUM, cex = 0.5)
    
    
  # Highlight PSA 77 & 183
    text(a[,1], a[,2], map_ppd$PSA_NUM, cex = 0.5)
    plot(st_geometry(subset(map_ppd, PSA_NUM == "077")),
         add = TRUE, border = "red")
    plot(st_geometry(subset(map_ppd, PSA_NUM == "183")),
         add = TRUE, border = "green")
    
```




### Exercise: Spatial joins & highlight areas with the most OIS
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Indicate coordinate system for latitude and longitude values
    ois <- st_as_sf(ois,
                     coords = c("lon", "lat"),
                     crs = st_crs("+init=epsg:4326")) #  World Geodetic System 1984 (used in GPS)

  # Match each row in "ois" to each PSA polygon
    lookup_psa <- st_join(ois, map_ppd[,"PSA_NUM"])
    lookup_psa[1:3, c("id", "date", "location", "PSA_NUM", "geometry")]
    
  # Find top 3 PSAs with the most OIs
    a <- rev(sort(table(lookup_psa$PSA)))
    a
    
  # Plot
    plot(st_geometry(map_ppd), axes = TRUE)
    i <- which(lookup_psa$PSA_NUM == names(a)[1])
    plot(st_geometry(lookup_psa[i,]), add = TRUE, col = "mediumpurple3", pch = 16)
    i <- which(lookup_psa$PSA_NUM == names(a)[2])
    plot(st_geometry(lookup_psa[i,]), add = TRUE, col = "steelblue1", pch = 16)
    i <- which(lookup_psa$PSA_NUM == names(a)[3])
    plot(st_geometry(lookup_psa[i,]), add = TRUE, col = "skyblue1", pch = 16)

    
    
    
```



### Exercise: Tabulate OIS count by PSA and add color to the map
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # How many shootings in each PSA?
    a <- table(lookup_psa$PSA_NUM)
    
  # Merge shooting counts into "map_ppd" data
    i <- match(map_ppd$PSA_NUM, names(a))
    map_ppd$nShoot <- a[i]
    map_ppd[1:3,]
    
  # Historgram to view distribution of counts
    hist(a, xlab = "Number of OIS", ylab = "Numbers of PSAs", main = "")
    
  # OIS counts into categories
    a <- cut(map_ppd$nShoot,
             breaks = c(0,1,5,10,15,20,25,30),
             right = FALSE)

  # Heat color for the map
    col <- rev(heat.colors(7,1))
    
  # Assign colors to each PSA & add PSA labels
    plot(st_geometry(map_ppd), col = col[as.numeric(a)], border = NA)
    a <- st_coordinates((st_centroid(map_ppd)))
    text(a[,1], a[,2], map_ppd$nShoot, cex = 0.5)
    
```











