---
title: "Working with Geography: Los Angeles Gang Injunction Maps and Crime"
author: "Pauline I. Alvarado"
date: "5/25/2019"
output: pdf_document
---

This coding exercise was from [University of Pennsylvania's Criminal Justice Data Science Course](https://github.com/gregridgeway/R4crim>) taught by Dr. Gregory Ridgeway. Script has been modified from course examples to fit personal notation conventions. California city shapes, California Census tracts, and Los Angeles County Rods were acquired from the [Census](https://www.census.gov/cgi-bin/geo/shapefiles/index.php) website. Los Angeles Safety Zones shapefiles were obtained by the Dr. Gregory K. Ridgeway from the Los Angeles City Attorney's Office. The 2014 crime data was taken from the [Los Angeles Open Data website](https://www.census.gov/cgi-bin/geo/shapefiles/index.php).  

Key skills: Managing Spatial Data, Manipulating Spatial Object, Accessing Data Through JSON, Merging ACS/LAPD/Gang Injunction Data/Census TIGER files


# Exploring Los Angeles Gang Injunction/Safety Zone Maps

## Data Wrangling 

### Load map and packages
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}
  # Load packages
    library(sf)
    library(lubridate)

  # Load Map
    map_sz <- st_read("allinjunctions/allinjunctions.shp")

```


### Plot the geometry of the shapefile
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  plot(st_geometry(map_sz))
  axis(1); axis(2); box() #add x and y axis for scale

```


### View the map projection of the shapefile
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  st_crs(map_sz)

```


### Examine the data attached to each polygon
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  map_sz[1:3,]

```


### Extract coordinates of the first injunction
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  st_coordinates(map_sz[1,])

```


### Clean up dates
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE, results='hold'}

  map_sz$Pre_Date <- as.character(map_sz$Pre_Date)
  map_sz$Pre_Date <- gsub("Jne", "June", map_sz$Pre_Date)
  map_sz$Pre_Date <- gsub("Sept\\.", "September", map_sz$Pre_Date)
  map_sz$Pre_Date <- mdy(map_sz$Pre_Date)
  map_sz$Perm_Date <- as.character(map_sz$Perm_Date)
  map_sz$Perm_Date <- gsub("Sept\\.", "September ", map_sz$Perm_Date)
  map_sz$Perm_Date <- mdy(map_sz$Perm_Date)
  map_sz$startDate <- pmin(map_sz$Pre_Date, map_sz$Perm_Date, na.rm = TRUE)

```


### Combine overlapping injuctions into one polygon
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Combine polygons  
    map_sz_union <- st_union(map_sz)
  
  # Check map units
    st_crs(map_sz_union)$units

```

### Plot combined polygon with a buffer zone that is within 500 feet of an injunction
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Create buffer zone
    map_sz_500 <- st_buffer(map_sz_union, dist=500)

  # Plot buffer zone
    plot(st_geometry(map_sz_500)) 
  
  # "Fill in" the combined injunction polygon in red
    plot(st_geometry(map_sz_union), col="red", border=NA, add=TRUE)
  
```


## Create MS-13 Injunction Shapefiles
Double-check the LA City Attorney's Office [website](http://freepdfhosting.com/0d363a8b29.pdf) with the correct injunction posted on its website.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Subset MS-13 data
    map_ms13 <- subset(map_sz, case_no=="BC311766")
  
  # Combine polygons
    a <- st_union(map_ms13)
  
  # Add back data
    map_ms13 <- st_sf(map_ms13[1,c("NAME","case_no","Safety_Zn","gang_name","startDate")],
                     geometry=a)
  
  # Plot MS-13 gang injunction safety zone and add 500ft buffer
    map_ms13_buff <- st_difference(st_geometry(st_buffer(map_ms13, dist=500)),
                               st_geometry(map_ms13))
    plot(st_geometry(map_ms13_buff), col="green")

```


# Merge LA City Gang Injunction and US Census Tract (TIGER) Shapefiles

## Extract outline of the city of Los Angeles from California Shapefile
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Load TIGER shapefiles
    map_ca <- st_read("tl_2014_06_place/tl_2014_06_place.shp") # california state polygon
    map_tract <- st_read("tl_2014_06_tract/tl_2014_06_tract.shp") # california census tracts polygon

  # Extract LA city outline from CA shapefile
    map_la <- subset(map_ca, NAMELSAD=="Los Angeles city")
    
  # Plot LA city polygon
    plot(st_geometry(map_la))
         
```


## Coordinate map projections for each shapefile
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  # Check projections
    st_crs(map_la) # longitude/latitude
    st_crs(map_tract) # longitude/latitude
    st_crs(map_sz) # lambert conformal conic
    
  # Turn map projections into LCC projection
    map_tract <- st_transform(map_tract, crs=st_crs(map_sz))
    map_la   <- st_transform(map_la,   crs=st_crs(map_sz))
```


## Merge LA City and Census Tract shapefiles

### Determine which census tract intersects with LA map
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # List of census tracts that intersect with map_la (only 1)
    a <- st_intersects(map_la, map_tract)

  # Create new column in map_tract to indicate if it's in LA
    map_tract$inLA <- FALSE
    map_tract$inLA[a[[1]]] <- TRUE
  
```
  
  
### Create LA city shapefile with census tracts
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
 
  map_la_tract <- subset(map_tract, inLA)
  plot(st_geometry(map_la_tract))
  plot(st_geometry(map_la), add=TRUE, border="red", lwd=3)

```


## Merge LA map with census tracts with MS-13 injunction map
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Create a TRUE / FALSE indicate of whether a tract is in MS-13
    st_intersects(map_ms13, map_la_tract)
    map_la_tract$inMS13 <- FALSE
    map_la_tract$inMS13[st_intersects(map_ms13, map_la_tract)[[1]]] <- TRUE
  
  # Merge shape files
    plot(st_geometry(subset(map_la_tract, inMS13)))
    plot(st_geometry(map_ms13), border="red", lwd=3, add=TRUE)
    
```


### Remove census tracts not in the MS-13 safety zone
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

    i <- st_intersects(st_buffer(map_ms13, dist = -200), map_la_tract)[[1]]
    map_la_tract$inMS13 <- FALSE
    map_la_tract$inMS13[i] <- TRUE
  
  # Merge shape files
    plot(st_geometry(subset(map_la_tract, inMS13)))
    plot(st_geometry(map_ms13), border="red", lwd=3, add=TRUE)
    
```


# Merge in Demographic Data from the American Community Survey

## Access ACS data via Javascript Object Notation (JSON)
For each census tract in LA County, extract total population (B03002001), number of non-Hispanic white residents (B03002003), non-Hispanic black residents (B03002004), and Hispanic residents (B03002012).
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Load relevant packages
    library(jsonlite)

  # Total Number of People by Specific Census Tracts
    acs_race <- fromJSON("https://api.census.gov/data/2017/acs/acs5?get=B03002_001E,B03002_003E,B03002_004E,B03002_012E&for=tract:*&in=state:06+county:037")
    
  # Convert into data frame
    a <- data.frame(acs_race[-1,], stringsAsFactors = FALSE)
    names(a) <- acs_race[1,]
    names(a)[1:4] <- c("total", "white", "black", "hisp")
    acs_race <- a
    for(i in c("total", "white", "black", "hisp"))
        acs_race[[i]] <- as.numeric(acs_race[[i]])
    
  # Add column of resident of other race groups
    acs_race$other <- with(acs_race, total-white-black-hisp)
 
```


## Add race information to MS-13 injunction data
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Match tract IDs and merge in percent hispanic
    i <- match(map_la_tract$TRACTCE, acs_race$tract)
    map_la_tract$pctHisp <- with(acs_race[i,], ifelse(total > 0 & !is.na(hisp), hisp/total, 0))
 
```


## Plot and format of percent hispanic residents within MS-13 safety zone by census tract
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Format shading by percentage
    col <- with(map_la_tract, gray(pctHisp[inMS13]))
    plot(st_geometry(subset(map_la_tract, inMS13)), col = col)
    plot(st_geometry(map_ms13), border = "red", lwd = 3, add = TRUE)
    
  # Text overlay
    labs <- with(map_la_tract, paste0(round(100*pctHisp[inMS13]), "%"))
    text(st_coordinates(st_centroid(subset(map_la_tract, inMS13))),
         labels = labs,
         cex = 0.5)
 
```


# Working with Point Data Using LA Crime Data

## Data Management of Data Taken from LA Open Data Portal
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Load file
    crime_data<- read.csv("LAPD_Crime_and_Collision_Raw_Data_-_2014.csv.gz",
                      as.is=TRUE)

  # Locate any geospatial variables
    names(crime_data)
    
  # Extract latitude and longitude information
    a <- gsub("[()]", "", crime_data$Location.1)
    crime_data$lat <- as.numeric(gsub(",.*", "", a))
    crime_data$lon <- as.numeric(gsub("[^,]*,", "", a))
    crime_data$Location.1 <- NULL
    crime_data <- subset(crime_data, !is.na(lat))
 
```


### Convert data frame into a simple features spatial object 
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Set to latitude/longitude coordinate system
    crime_data <- st_as_sf(crime_data, 
                          coords = c("lon","lat"),
                          crs = 4326) 

  # Set to the same coordinate system as the MS-13 map
    crime_data <- st_transform(crime_data, st_crs(map_ms13))

```


## Identify crime data occurring within 1 mile of MS-13 injunction.

### Subset LAPD areas (1,2,3,6,7, and 20) that intersected with MS-13 dafety zone.
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  crime_ms13 <- subset(crime_data, AREA %in% c(1,2,3,6,7,11,20))

```


### Create variable to code location around MS-13 injuction (within safety zone, buffer, and outside)
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # create variable to label crime's location
    crime_ms13$place1 <- "outside"
    i <- st_intersects(map_ms13, crime_ms13)[[1]]
    crime_ms13$place1[i] <- "SZ"
    i <- st_intersects(st_difference(st_buffer(map_ms13, dist = 5280), map_ms13), crime_ms13)[[1]]
    crime_ms13$place1[i] <- "buffer"
    
  # plot to see if correctly labeled
    plot(st_geometry(st_buffer(map_ms13, dist = 5280)))
    plot(st_geometry(map_ms13), border = "red", lwd = 3, add = TRUE)
    plot(st_geometry(subset(crime_ms13, place1 == "SZ")), pch = ".", col = "red", add = TRUE)
    plot(st_geometry(subset(crime_ms13, place1 == "buffer")), pch = ".", col = "blue", add = TRUE)
    plot(st_geometry(subset(crime_ms13, place1 == "outside")), pch = ".", col = "green", add = TRUE)
    
  # combine geometries of three polygons
    map_a <- c(st_geometry(map_ms13),
               st_geometry(st_difference(st_buffer(map_ms13, dist = 5280), map_ms13)),
               st_geometry(st_difference(st_buffer(map_ms13, dist = 80*5280), st_buffer(map_ms13, dist = 5280))))
    
  # create sf object
    map_a <- st_sf(place2 = c("SZ", "buffer", "outside"), geom = map_a)
    crime_ms13 <- st_join(crime_ms13, map_a)
      
``` 


# Creating New Geographic Objects

## Extract the Southern Component of the MS-13 Map
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # get sf object information
    is(st_geometry(map_ms13))

  # break apart multipolygon object into polygons
    a <- st_cast(map_ms13, "POLYGON")
    
  # store southern polygon
    ms13_south <- a[1,]
    plot(st_geometry(ms13_south))
```


# Overlaying a Street Map

## Merge TIGER maps with streets with injunction map
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # load TIGER line file
    la_street <- st_read("tl_2014_06037_roads/tl_2014_06037_roads.shp")

  # use the same projection as a the injunction map
    la_street <- st_transform(la_street, st_crs(ms13_south))
    
  # extract only the streets that instersect with the MS-13 south safety zone
    la_street$inSZ <- FALSE
    i <- st_intersects(ms13_south, la_street)[[1]]
    la_street$inSZ[i] <- TRUE
    ms13_south_st <- subset(la_street, inSZ)
    
  # plot
    plot(st_geometry(ms13_south_st))
    plot(st_geometry(ms13_south), border = "red", lwd = 3, add = TRUE)
    
```


## Add street names 

### Function on where and what angle to place street names
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # extract coords for every street segment
    a <- lapply(st_geometry(ms13_south_st), st_coordinates)
                
  # get x,y, and angle for each street segment
   labs <- sapply(a, function(coord)
   {
      # which parts of the street are inside MS13 safety zone
      i <- which((coord[,"X"] > st_bbox(ms13_south_st)["xmin"]) &
                 (coord[,"X"] < st_bbox(ms13_south_st)["xmax"]) &
                 (coord[,"Y"] > st_bbox(ms13_south_st)["ymin"]) &
                 (coord[,"Y"] < st_bbox(ms13_south_st)["ymax"]))
      # don't select the last one, too close to the edge
      i <- setdiff(i, nrow(coord))
      # if none are in bounding box just use the first coordinate
      if(length(i)==0) i <- 1
      # randomly choose a point on the street for the label
      i <- sample(i, size=1)
      # compute the slope  of the street, change in y/change in x
      streetSlope <- (coord[i+1,2]-coord[i,2]) / (coord[i+1,1]-coord[i,1])
      # compute the angle of the slope with the arc-tangent
      angle <- atan(streetSlope)
      # atan() returns radians, convert to degrees
      angle <- 180*angle/pi
      # round to the nearest 10
      angle <- round(angle, -1)
      # would rather not have labels that are upside down
      angle <- ifelse(angle < -90,  180+angle, angle)
      angle <- ifelse(angle >  90, -180+angle, angle)
      
      return(c(x=coord[i,1], y=coord[i,2], angle=angle))
   })
  
  # transpose results and get data frame
   labs <- data.frame(t(labs))
   
```


### Plot
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  plot(st_geometry(ms13_south), border = "red", lwd = 1)
  plot(st_geometry(ms13_south_st), add = TRUE)
  plot(st_geometry(ms13_south), border = "red", lwd = 3, add = TRUE)
  
  for(i in 1:nrow(labs))
  {
     text(labs$x[i], labs$y[i],
          ms13_south_st$FULLNAME[i],
          srt = labs$angle[i],         # srt = string rotation
          cex = 0.6)                   # cex = character expansion
  }
  
```


# Geospatial Analysis of Crime in the MS-13 Safety Zone

## CRIME COUNT within 100 feet of Wilshire Blvd

### Highlight Wilshire Ave on the map
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  plot(st_geometry(ms13_south), border = "red", lwd = 1)
  plot(st_geometry(ms13_south_st), add = TRUE)
  plot(st_geometry(ms13_south), border = "red", lwd = 3, add = TRUE)
  
  for(i in 1:nrow(labs))
  {
     text(labs$x[i], labs$y[i],
          ms13_south_st$FULLNAME[i],
          srt = labs$angle[i],         # srt = string rotation
          cex = 0.6)                   # cex = character expansion
  }
  
  wilshire_st <- subset(ms13_south_st, FULLNAME == "Wilshire Blvd")

  plot(st_geometry(wilshire_st), col = "green", lwd = 3, add = TRUE)
  
```


### Create 100ft buffer around Wilshire, exclude outside of MS-13 safety zone
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # create a 100-foot buffer, but only the part that is in the ms13 safety zone
    wilshire_buff <- st_intersection(st_geometry(st_buffer(wilshire_st, dist = 100)),
                                      st_geometry(ms13_south))

  # subset crime data within that buffer
    i <- st_intersects(wilshire_buff, crime_ms13)[[1]]
    crime_ms13$inWilbuf <- FALSE
    crime_ms13$inWilbuf[i] <- TRUE
    
    plot(st_geometry(ms13_south), border="red", lwd = 1)
    plot(st_geometry(wilshire_buff), add = TRUE, border = "green")
    plot(st_geometry(ms13_south_st), add = TRUE)
    plot(st_geometry(ms13_south), border = "red", lwd=3, add = TRUE)
    
    for(i in 1:nrow(labs))
    {
       text(labs$x[i], labs$y[i],
            ms13_south_st$FULLNAME[i],
            srt = labs$angle[i],         # srt = string rotation
            cex = 0.6)                   # cex = character expansion
    }
    
    plot(st_geometry(subset(crime_ms13, inWilbuf)), col = "blue", add = TRUE, pch = 16, cex = 0.5)
    
```


## COMMON CRIMES withn 100 feet of Wilshire Blvd
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  with(crime_ms13, rev(sort(table(Crm.Cd.Desc[inWilbuf])))[1:5]) 

```


## Streets with MAX INCIDENTS of crime

### Subset crime data for only MS-13 South
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  i <- st_intersects(ms13_south, crime_ms13)[[1]]
  crime_ms13_south <- crime_ms13[i,]
  plot(st_geometry(ms13_south), border = "red", lwd = 1)
  plot(st_geometry(ms13_south_st), add = TRUE)
  plot(st_geometry(ms13_south), border = "red", lwd = 3, add = TRUE)
  plot(st_geometry(crime_ms13_south),
       add = TRUE, col = "blue",pch = 16, cex = 0.5)
  
```


### Compute the distance for each point to the closest street in `ms13_south_st`
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  d <- st_distance(crime_ms13_south, ms13_south_st)
  dim(d) # row for each crime, column for each street

  # for each row (crime) find out which column (street)
    iClose <- apply(d, 1, which.min)

  # for the first crime check that the original address is similar to closest street
    crime_ms13_south[1,]
    ms13_south_st[iClose[1],]
    plot(st_geometry(ms13_south), border = "red", lwd = 1)
    plot(st_geometry(ms13_south_st), add = TRUE)
    plot(st_geometry(ms13_south), border = "red", lwd = 3, add = TRUE)
    
    for(i in 1:nrow(labs))
    {
       text(labs$x[i], labs$y[i],
            ms13_south_st$FULLNAME[i],
            srt = labs$angle[i],
            cex = 0.6)
    }
    
    plot(st_geometry(crime_ms13_south[1,]),
         add = TRUE, col = "red", pch = 16, cex = 2)
```
  

### Streets with the most incidents using distance calculation
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
  
  a <- table(ms13_south_st$FULLNAME[iClose])
  rev(sort(a))[1:10]

```

## CRIME COUNT within 500ft of a Metrorail station

### Map stations (Wilshire at Western Ave, S Normandie Ave, S Vermont Ave, Alvarado) and buffer zone
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  map_metro <- st_intersection(subset(la_street, FULLNAME %in% 
                                c("S Western Ave", "S Normandie Ave", 
                                  "S Vermont Ave", "S Alvarado St")),
                      subset(la_street, FULLNAME == "Wilshire Blvd"))

  plot(st_geometry(ms13_south), border = "red", lwd = 1)
  plot(st_geometry(ms13_south_st), add = TRUE)
  plot(st_geometry(ms13_south), border = "red", lwd = 3, add = TRUE)
  plot(st_geometry(map_metro), col = "purple", add = TRUE, pch = 16, cex = 2)
  
  plot(st_geometry(st_buffer(map_metro, dist = 500)), add = TRUE, border = "purple")

```


### Crime counts for each station
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # Each station
  i <- st_intersects(st_buffer(map_metro, dist = 500), crime_data)
  sapply(i, length)

  # Overall
  length(unlist(i))

```


## CRIME COUNT within 500ft of the RFK Community School

### Plot the school area and buffer
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  # plot the intersection points
    a <- st_intersection(subset(la_street, FULLNAME %in% c("S Mariposa Ave", "S Catalina St")),
                       subset(la_street, FULLNAME %in% c("W 8th St", "Wilshire Blvd")))

    plot(st_geometry(ms13_south), border = "red", lwd = 1)
    plot(st_geometry(ms13_south_st), add = TRUE)
    plot(st_geometry(ms13_south), border = "red", lwd = 3, add = TRUE)
    
    a <- a[-which.min(st_coordinates(a)[,1]),] #drop Maripose/Wilshire intersection
    
    plot(st_geometry(a), col = "orange", add = TRUE, pch = 16, cex = 2)
    
  # compute the convex hull of all 4 points
    map_rfk <- st_convex_hull(st_union(a))
    plot(st_geometry(map_rfk), border = "purple", add = TRUE, lwd = 3)
    
```


### Total crime count within 500ft of RFK School
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

  length(st_intersects(st_buffer(map_rfk, dist = 500), crime_ms13_south))

```



# Keyhole Markup Language (KML) File for Google Maps
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

 a <- st_transform(map_ms13, crs = 4326)
  
  st_write(a,
           dsn = "ms13.kml",
           layer = "ms13",
           driver = "KML",
           delete_dsn = TRUE)

```